Nonparametric Clustering from Energy Statistics

### 2017-08-11 qcqp_clustering/draft/energy_clustering.tex
finished draft
corrected typos, included reference

%\usepackage{nicefrac}       % compact symbols for 1/2, etc.
%\usepackage{microtype}      % microtypography
\usepackage{amsmath,amssymb,amsfonts,amsthm,amscd,bm,bbm}
%\usepackage{mathtools}
%\usepackage{printlen}
%\usepackage[inline]{enumitem}
%\usepackage{cite}
%\usepackage{bbm}

@@

\begin{abstract}
Energy statistics 
was proposed by 
Sz\' ekely in the 80's
inspired by the Newtonian gravitational potential from classical mechanics,
and it provides a nonparametric test for equality of distributions.
It was generalized to 
metric spaces of strong negative type,
and more recently a connection with reproducing kernel Hilbert spaces (RKHS)
was proposed.
Although  
extensively used in statistics, it
has not been previously incorporated in machine learning problems.
Here we consider the problem of clustering data from
an energy statistics theory perspective.
We provide a precise mathematical formulation yielding
a quadratically constrained quadratic program (QCQP), which we show
to be equivalent
to kernel $k$-means optimization problem.
Thus, our results imply a first principles derivation of 
kernel $k$-means from energy statistics.
Moreover, we also consider a weighted version of energy statistics 
applied to clustering, which is related to 
graph partitioning problems.
To find local optimizers of such QCQP we consider an iterative
algorithm based on Hartigan's method, which in this case
has the same computational cost 
as kernel $k$-means algorithm but usually with better clustering
performance. We provide carefully designed numerical experiments
showing the superiority of this method compared to
kernel $k$-means, standard $k$-means, and gaussian mixture models.
\end{abstract}

@@

Although clustering from energy statistics was considered
in \cite{Kgroups}, the precise optimization problem behind this approach
remains elusive, as well as the connection to kernel methods.
The main theoretical contribution of this paper is to fill this gap.
Since the statistical potential energy is minimum when
distributions are equal, the principle 
principle behind clustering is to maximize the statistical energy, 
enforcing probability distributions associated to each cluster
to be different from one another.
We provide a precise mathematical formulation to this
statement, showing that there is only one possible optimization problem
consistent with energy statistics. Moreover, we show that this is
a quadratically constrained
quadratic program (QCQP) in the associated RKHS. Our results immediately 
establish the connection 
to kernel methods used in machine learning, 
by showing that this QCQP is equivalent
to kernel $k$-means optimization problem. 
The equivalence between kernel $k$-means, 
spectral clustering, and graph partitioning problems was already established
in \cite{Dhillon,Dhillon2}. We show how these connections arise
from a weighted version of energy statistics.

Our algorithmic contribution is to use Hartigan's method
\cite{Hartigan} to find local solutions of the above mentioned QCQP.
This approach was also proposed in \cite{Kgroups}. In this case,
the resulting algorithm has the same
time complexity as kernel $k$-means algorithm, however, the numerical 
results provide compelling evidence that this method
is more accurate and robust than kernel $k$-means. This evidence
is also supported by general theoretical results \cite{Telgarsky,Slonin}.
Moreover, in our experiments we put in
in evidence the nonparametric aspect of energy
statistics based clustering, which provides a default kernel, showing that
it is able to perform accurately on data coming very different 
distributions, contrary to $k$-means and GMM for instance.

@@

Finally, let us recall the main formulas from energy statistics
for the test statistic of equality of distributions \cite{Szkely2013}. 
Assume we have data $\mathbb{X} = \{ x_1,\dotsc, x_n \}$ where
$x_i \in \mathcal{X}$ and $\mathcal{X}$ is a space of negative type.
Consider a disjoint partition $\mathbb{X} = \bigcup_{j=1}^k \C_j$, with
$\C_i \cap \C_j = \emptyset$.
Each expectation in 
\eqref{eq:energy3}
can be computed 
through the function
\begin{equation}
\label{eq:g_def}
g (\C_i, \C_j) \equiv 
\dfrac{1}{n_i n_j}
\sum_{x \in \C_i} 
\sum_{y \in \C_j} \rho(x, y)
\end{equation}
where $n_i = |\C_i|$ is the number of elements in 
$\C_i$. 
The \emph{within energy dispersion} is defined by
\begin{equation}
\label{eq:within}
W \equiv
\sum_{j=1}^{k} \dfrac{n_j}{2} g(\C_j, \C_j)
\end{equation}
and the \emph{between-sample energy statistic} is defined by
\begin{equation}
\label{eq:between}
S \equiv
\sum_{1 \le  i < j \le k } \dfrac{n_i n_{j}}{2 n} \left[
2 g(\C_i, \C_j) - 
g(\C_i, \C_i) - 
g(\C_j, \C_j)
\right]
\end{equation}
where $n = \sum_{j=1}^k n_j$.
Given a set of distributions
$\{ P_j\}_{j=1}^k$, where $x \in \C_j$ if and only if $x \sim P_j$, 
the quantity $S$ provides
a \emph{nonparametric test statistic} for equality of distributions
\cite{Szkely2013}.
When the sample size is large enough, $n\to \infty$,
under the null hypothesis $H_0: P_1=P_2=\dotsm=P_k$ we have that
$S\to 0$, 
and under
the alternative hypothesis $H_1: P_i \ne P_j$ for at least two $i\ne j$, 
we have that $S \to \infty$.
This test is nonparametric in the sense that it does not make any assumptions
about the distributions $P_j$.

@@

Due to the test statistic \eqref{eq:between} for equality of distributions,
the obvious
criterion for clustering data is to 
maximize $S$, which makes 
each cluster as different
as possible from the other ones.
In other words, given a set of points coming from different probability
distributions, $S$ should attain a maximum when each point is correctly
classified as belonging to the cluster associated to its probability
distribution.
The following 
straightforward result
shows that maximizing \eqref{eq:between} is, however, equivalent to minimizing
\eqref{eq:within} which has a more convenient form.

@@

For a given $k$, the clustering problem amounts to
finding the best partition of the data by solving~\eqref{eq:minimize}.
Notice that this is a hard clustering problem, as partitions
are disjoint. The problem \eqref{eq:minimize} based on
energy statistics was already proposed in \cite{Kgroups}. However, it is
important to note that this is equivalent to maximizing \eqref{eq:between}
which is the test statistic for equality of distributions. In the following
we show what is the explicit optimization problem behind \eqref{eq:minimize}.

Let us formulate \eqref{eq:minimize} in the corresponding
RKHS. Based on \eqref{eq:kernel_semimetric} and \eqref{eq:gen_kernel}, 
assume that the kernel $\kk: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$ 
generates $\rho$.  Let us define  the Gram matrix
\begin{equation}
\label{eq:kernel_matrix}
G \equiv \begin{pmatrix}
\kk(x_1,x_1) & \kk(x_1,x_2) & \dotsm & \kk(x_1,x_n) \\
\kk(x_2,x_1) & \kk(x_2,x_2) & \dotsm & \kk(x_2,x_n) \\
\vdots & \vdots & \ddots  & \vdots \\
\kk(x_n,x_1) & \kk(x_n,x_2) & \dotsm & \kk(x_n,x_n) 
\end{pmatrix} .
\end{equation}
Let $Z \in \{ 0,1 \}^{n\times k}$ be the label matrix, 
with only one nonvanishing entry per row, 
indicating to which cluster (column)
each point (row) belongs to. This matrix satisfy
$Z^\top Z = D$ where the diagonal matrix 
$D = \diag( n_1,\dotsc, n_k )$  contains
the number of points in each cluster. Let us also introduce the rescaled
matrix  $Y \equiv Z D^{-1/2}$. In component form they are given by
\begin{equation}
\label{eq:label_matrix}
Z_{ij} \equiv \begin{cases}
1 & \mbox{if $x_i \in \C_j$ } \\
0 & \mbox{otherwise}
\end{cases} \qquad
\Zt_{ij} \equiv \begin{cases}
\tfrac{1}{\sqrt{n_j}} & \mbox{if $x_i \in \C_j$ } \\
0 & \mbox{otherwise}
\end{cases} .
\end{equation}
Throughout the paper, we use the notation $M_{i\bullet}$ to denote
the $i$th row of a matrix $M$, and $M_{\bullet j}$ denotes its $j$th column.
Our next result shows the optimization problem \eqref{eq:minimize}
is NP-hard since
it is a quadratically constrained quadratic program (QCQP).

@@

Based on Proposition~\ref{th:qcqp2}, to group data
$\mathbb{X} = \{ x_1,\dotsc,x_n \}$
into  $k$ clusters, we first compute the Gram matrix
$G$ and then 
solve the optimization problem \eqref{eq:qcqp2} for $\Zt \in
\mathbb{R}^{n\times k}$. The $i$th row
of $\Zt$ will contain a single nonzero element in some $j$th column,
indicating that $x_i \in \C_j$. 
Problem \eqref{eq:qcqp2} is NP-hard and there
are few methods
available to solve it directly,
which is computational prohibitive even for small datasets.
However, one can find approximate solutions by relaxing some 
of the constraints, or obtaining a relaxed SDP version of it.
For instance, the relaxed problem
\begin{equation}
\max_{Y} \Tr \left( Y^\top G \, Y \right) \quad \mbox{s.t. $Y^\top Y = I$}
\end{equation}
has a well-known closed form solution $Y^\star = U R$, where the
columns of $U$ contain the leading $k$ eigenvectors of $G$ corresponding
to the $k$ largest eigenvalues $\lambda_1\ge \lambda_2\ge\dotsc\ge\lambda_k$, 
and
$R \in \mathbb{R}^{k\times k}$ is an arbitrary orthogonal matrix. 
The resulting
optimal objective function is given by
$\max \Tr \left( {Y^\star}^\top G \, Y^\star \right)  = 
\sum_{i=1}^k \lambda_i$. One might then normalize and threshold the rows
of $Y^\star$, or better, following \cite{NgJordan} one can normalize the
rows of $Y^\star$ and apply standard $k$-means on this matrix where each
row is considered as a data point.
This is the procedure done in spectral clustering on a graph Laplacian
matrix defined by a similarity matrix.
However, computing eigenvectors of a very large matrix
can be very expensive, and usually iterative methods are preferred. Moreover,
this procedure is only guaranteed to converge if the normalized $Y^\star$ is
positive semidefinite.
In Section~\ref{sec:algo} we will propose an iterative method to directly 
find approximate solutions to \eqref{eq:qcqp2} that is guaranteed to
converge and does not require that the similarities $\kk(x_i,x_j)$ be
nonnegative.

The clustering problem \eqref{eq:qcqp2} based on energy statistics
is valid for data living in an \emph{arbitrary} space of negative type, where
a semimetric $\rho$, and thus the kernel \eqref{eq:kernel_semimetric}, are
assumed to be known. This clustering method is
\emph{nonparametric} since it does not make any assumptions
about  the distribution of the data, 
contrary to $k$-means and gaussian mixture models (GMM), for example.
%Moreover, this approach \emph{does not} require the concept of the 
%\emph{cluster mean},
%which can be ill-defined for some types of data, such as images for
%instance. Since no mean is involved, the method should also be 
%robust to outliers.
%This method is quite general and makes very few
%assumptions about the data. 
If one uses the standard energy distance
defined in \eqref{eq:energy}, 
then $\rho(x,y) = \| x - y\|$ and this fixes the kernel
through \eqref{eq:kernel_semimetric}. In the same way, 
for data living in a more general
metric space $(\mathcal{X}, \rho)$ the corresponding semimetric $\rho$ fixes
the kernel.
In practice, however,
the clustering quality strongly depend on the choice of a suitable
$\rho$ which measures the similarity between data points,
and is equivalent to choosing an appropriate kernel.
The energy distance \eqref{eq:energy2} fixes a family of choices
$\rho(x,y) = \| x-y\|^\alpha$ with $0 < \alpha\le 2$. If prior information
is available one may choose a more appropriate $\rho$.

\subsection*{Relation to Kernel $\bm{k}$-Means}
One may wonder how energy statistics clustering 
relates to the well-known kernel $k$-means problem%
\footnote{When we refer to kernel $k$-means problem we mean specifically 
the optimization problem \eqref{eq:kernel_kmeans}, which should not be 
confused with kernel $k$-means algorithm that is just one possible recipe 
to solve \eqref{eq:kernel_kmeans}.
} 
which is extensively used in machine learning.
%and was proposed as a kernelized version of $k$-means.
For a positive semidefinite Gram matrix $G$, as defined in
\eqref{eq:kernel_matrix},
there exists a map
$\varphi: \mathcal{X} \to \HH_\kk$ such that
$\kk(x,y) = \varphi(x)^\top \varphi(y)$. The kernel $k$-means optimization
problem,
in the feature space,
is defined by
\begin{equation}
\label{eq:kernel_kmeans}
\min_{\C_1,\dotsc,\C_k}\bigg\{ 
J(\C_1,\dots,\C_k) \equiv  \sum_{j=1}^k
\sum_{x \in \C_j} \| \varphi(x) - \varphi(\mu_j) \|^2
\bigg\},
\end{equation}
where $\mu_j = \tfrac{1}{n_j} \sum_{x \in \C_j} x$ is the  mean of cluster
$\C_j$ in the ambient space. Notice that the above objective function
is strongly tied to the idea of minimizing distances between points
and cluster centers, which arises from $k$-means objective function.
It was shown \cite{Dhillon2,Dhillon} 
that problem \eqref{eq:kernel_kmeans} 
is equivalent to a QCQP in the same form as
\eqref{eq:qcqp2}. The next result makes this explicit, showing that
\eqref{eq:minimize} and \eqref{eq:kernel_kmeans} are actually equivalent.

@@

\begin{proposition}
\label{th:qcqp3}
The weighted version of energy statistics clustering given by
problem \eqref{eq:minimize2} is equivalent to
\begin{equation}
\label{eq:qcqp3}
\max_H \Tr \left\{ H^\top (\mathcal{W}^{1/2} G \mathcal{W}^{1/2}) H  \right\}
\qquad \mbox{s.t. $H \ge 0$, $H^\top H = I$, $H H^\top \om =
\om$,}
\end{equation}
where $G$ is the Gram matrix \eqref{eq:kernel_matrix} and the other quantities
are defined in 
\eqref{eq:weighted_matrices}.
\end{proposition}
\begin{proof}
Replacing \eqref{eq:gen_kernel} and eliminating the global terms which 
do not contribute, the optimization problem \eqref{eq:minimize2}
becomes 
\begin{equation}
\max_{\C_1,\dotsc,\C_k} \sum_{j=1}^k \dfrac{1}{s_j}
\sum_{x\in\C_j}\sum_{y\in\C_j} w(x)w(y) \kk(x,y) . 
\end{equation}
This 
objective function can be written as
\begin{equation}
\begin{split}
\sum_{j=1}^k \dfrac{1}{s_j} 
\sum_{p=1}^n \sum_{q=1}^n 
w_p w_q Z_{pj} Z_{qj} G_{pq} &= 
\sum_{j=1}^k 
\sum_{p=1}^n \sum_{q=1}^n 
\dfrac{Z^\top_{jp}\sqrt{w_p}}{\sqrt{s_j}} w_p^{1/2} G_{pq} w_q^{1/2} 
\dfrac{\sqrt{w_q} Z_{qj}}{\sqrt{s_j}} \\
&= 
\sum_{j=1}^k \left(H^\top \mathcal{W}^{1/2} G \mathcal{W}^{1/2} H\right)_{jj}
\\
&= \Tr\left( H^\top \mathcal{W}^{1/2} G \mathcal{W}^{1/2} H  \right).
\end{split}
\end{equation}
To obtain the constraints, note that $H_{ij} \ge 0$ by definition, and
\begin{equation}
(H^\top H)_{ij} = \sum_{\ell=1}^n 
Y_{\ell i} \mathcal{W}_{\ell \ell} Y_{\ell j } = 
\dfrac{1}{\sqrt{s_i}\sqrt{s_j}} \sum_{\ell=1}^n w_\ell Z_{\ell i} Z_{\ell j}
= \dfrac{\delta_{ij}}{s_i} \sum_{\ell=1}^n w_\ell Z_{\ell i} = \delta_{ij},
\end{equation}
therefore $H^\top H = I$. This is a constraint on the rows of $H$.
To obtain a condition on the columns
of $H$ observe that
\begin{equation}
\left(H^\top H\right)_{pq} = \sqrt{w_p w_q}\sum_{j=1}^k \dfrac{Z_{pj}
Z_{qj}}{s_j} = \begin{cases}
\dfrac{\sqrt{w_p w_q}}{s_i} & \mbox{if both $x_p,x_q \in \C_i$} \\
0 & \mbox{otherwise}.
\end{cases}
\end{equation}
Therefore, $(H^\top H \mathcal{W}^{1/2})_{pq} = \sqrt{w_p} \, w_q s_i^{-1}$
if both points $x_p$ and $x_q$ belong to the same cluster, which
we denote by $\C_i$ for some $i=1,\dotsc,k$, and 
$(H^\top H \mathcal{W}^{1/2})_{pq} = 0 $ otherwise. Thus, the $p$th
line of this matrix is nonzero only on entries corresponding to points
that are in the same cluster as $x_p$. If we sum over the columns of this
line we obtain $\sqrt{w_p} s_i^{-1} \sum_{q=1}^n w_q Z_{qi} = \sqrt{w_p}$,
or equivalently $H H^\top \mathcal{W}^{1/2} \e = \mathcal{W}^{1/2} \e$,
which gives the constraint $H H^\top \om = \om$.
\end{proof}

@@

The relation between kernel $k$-means and graph partitioning problems
was already established \cite{Dhillon2,Dhillon}. We repeat a similar analysis
due to the relation of these problems to
energy statistics and RKHS, which may provide a different perspective.

>>>>

%\usepackage{nicefrac}       % compact symbols for 1/2, etc.
%\usepackage{microtype}      % microtypography
\usepackage{amsmath,amssymb,amsfonts,amsthm,amscd,bm,bbm}
%\usepackage{mathtools}
%\usepackage{printlen}
\usepackage[inline]{enumitem}
%\usepackage{cite}
%\usepackage{bbm}

@@

\begin{abstract}
Energy statistics 
was proposed by 
Sz\' ekely in the 80's
inspired by the Newtonian gravitational potential from classical mechanics,
and it provides a nonparametric test for equality of distributions.
It was generalized to 
metric spaces of strong negative type,
and more recently a connection with reproducing kernel Hilbert spaces (RKHS)
was proposed.
Although  
extensively used in statistics, it
has not been previously incorporated in machine learning problems.
Here we consider the problem of clustering data from
an energy statistics theory perspective.
We provide a precise mathematical formulation yielding
a quadratically constrained quadratic program (QCQP), which we show
to be equivalent
to kernel $k$-means optimization problem.
Thus, our results imply a first principles derivation of 
kernel $k$-means from energy statistics.
Moreover, we also consider a weighted version of energy statistics 
applied to clustering, which makes connections to 
graph partitioning problems.
To find local optimizers of such QCQP we consider an iterative
algorithm based on Hartigan's method, which in this case
has the same computational cost 
as kernel $k$-means algorithm but usually with better clustering
performance. We provide carefully designed numerical experiments
showing the superiority of this method compared to
kernel $k$-means, standard $k$-means, and gaussian mixture models.
\end{abstract}

@@

Although clustering from energy statistics was considered
in \cite{Kgroups}, the precise optimization problem behind this approach
remains elusive, as well as the connection to kernel methods.
The main theoretical contribution of this paper is to fill this gap.
Since the statistical potential energy is minimum when
distributions are equal, the
principle behind clustering is to maximize the statistical energy, 
enforcing probability distributions associated to each cluster
to be different from one another.
We provide a precise mathematical formulation to this
statement, showing that there is only one possible optimization problem
consistent with energy statistics. Moreover, we show that this is
a quadratically constrained
quadratic program (QCQP) in the associated RKHS. Our results immediately 
establish the connection 
to kernel methods used in machine learning, 
by showing that this QCQP is equivalent
to kernel $k$-means optimization problem. 
The equivalence between kernel $k$-means, 
spectral clustering, and graph partitioning problems was already established
in \cite{Dhillon,Dhillon2}. We show how these connections arise
from a weighted version of energy statistics.

Our algorithmic contribution is to use Hartigan's method
\cite{Hartigan} to find local solutions of the above mentioned QCQP.
This approach was also proposed in \cite{Kgroups}. In this case,
the resulting algorithm has the same
time complexity as kernel $k$-means algorithm, however, the numerical 
results provide compelling evidence that this method
is more accurate and robust than kernel $k$-means. This evidence
is also supported by general theoretical results \cite{Telgarsky,Slonin}.
Moreover, in our experiments we put in
evidence the nonparametric aspect of energy
statistics based clustering, which provides a default kernel, showing that
it is able to perform accurately on data coming very different 
distributions, contrary to $k$-means and GMM for instance.

@@

Finally, let us recall the main formulas from energy statistics
for the test statistic of equality of distributions \cite{Szkely2013}. 
Assume we have data $\mathbb{X} = \{ x_1,\dotsc, x_n \}$ where
$x_i \in \mathcal{X}$, and $\mathcal{X}$ is a space of negative type.
Consider a disjoint partition $\mathbb{X} = \bigcup_{j=1}^k \C_j$, with
$\C_i \cap \C_j = \emptyset$.
Each expectation in 
\eqref{eq:energy3}
can be computed 
through the function
\begin{equation}
\label{eq:g_def}
g (\C_i, \C_j) \equiv 
\dfrac{1}{n_i n_j}
\sum_{x \in \C_i} 
\sum_{y \in \C_j} \rho(x, y)
\end{equation}
where $n_i = |\C_i|$ is the number of elements in 
$\C_i$. 
The \emph{within energy dispersion} is defined by
\begin{equation}
\label{eq:within}
W \equiv
\sum_{j=1}^{k} \dfrac{n_j}{2} g(\C_j, \C_j)
\end{equation}
and the \emph{between-sample energy statistic} is defined by
\begin{equation}
\label{eq:between}
S \equiv
\sum_{1 \le  i < j \le k } \dfrac{n_i n_{j}}{2 n} \left[
2 g(\C_i, \C_j) - 
g(\C_i, \C_i) - 
g(\C_j, \C_j)
\right]
\end{equation}
where $n = \sum_{j=1}^k n_j$.
Given a set of distributions
$\{ P_j\}_{j=1}^k$, where $x \in \C_j$ if and only if $x \sim P_j$, 
the quantity $S$ provides
a \emph{nonparametric test statistic} for equality of distributions
\cite{Szkely2013}.
When the sample size is large enough, $n\to \infty$,
under the null hypothesis $H_0: P_1=P_2=\dotsm=P_k$ we have that
$S\to 0$, 
and under
the alternative hypothesis $H_1: P_i \ne P_j$ for at least two $i\ne j$, 
we have that $S \to \infty$.
This test is nonparametric in the sense that it does not make any assumptions
about the distributions $P_j$.

@@

Due to the test statistic \eqref{eq:between} for equality of distributions,
the obvious
criterion for clustering data is to 
maximize $S$ which makes 
each cluster as different
as possible from the other ones.
In other words, given a set of points coming from different probability
distributions, $S$ should attain a maximum when each point is correctly
classified as belonging to the cluster associated to its probability
distribution.
The following 
straightforward result
shows that maximizing \eqref{eq:between} is, however, equivalent to minimizing
\eqref{eq:within} which has a more convenient form.

@@

For a given $k$, the clustering problem amounts to
finding the best partition of the data by solving~\eqref{eq:minimize}.
Notice that this is a hard clustering problem as partitions
are disjoint. The problem \eqref{eq:minimize} based on
energy statistics was already proposed in \cite{Kgroups}. However, it is
important to note that this is equivalent to maximizing \eqref{eq:between}
which is the test statistic for equality of distributions. In the following
we show what is the explicit optimization problem behind \eqref{eq:minimize}.

Let us formulate \eqref{eq:minimize} in the corresponding
RKHS. Based on \eqref{eq:kernel_semimetric} and \eqref{eq:gen_kernel}, 
assume that the kernel $\kk: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$ 
generates $\rho$.  Let us define  the Gram matrix
\begin{equation}
\label{eq:kernel_matrix}
G \equiv \begin{pmatrix}
\kk(x_1,x_1) & \kk(x_1,x_2) & \dotsm & \kk(x_1,x_n) \\
\kk(x_2,x_1) & \kk(x_2,x_2) & \dotsm & \kk(x_2,x_n) \\
\vdots & \vdots & \ddots  & \vdots \\
\kk(x_n,x_1) & \kk(x_n,x_2) & \dotsm & \kk(x_n,x_n) 
\end{pmatrix} .
\end{equation}
Let $Z \in \{ 0,1 \}^{n\times k}$ be the label matrix, 
with only one nonvanishing entry per row, 
indicating to which cluster (column)
each point (row) belongs to. This matrix satisfy
$Z^\top Z = D$ where the diagonal matrix 
$D = \diag( n_1,\dotsc, n_k )$  contains
the number of points in each cluster. Let us also introduce the rescaled
matrix  $Y \equiv Z D^{-1/2}$. In component form they are given by
\begin{equation}
\label{eq:label_matrix}
Z_{ij} \equiv \begin{cases}
1 & \mbox{if $x_i \in \C_j$ } \\
0 & \mbox{otherwise}
\end{cases} \qquad
\Zt_{ij} \equiv \begin{cases}
\tfrac{1}{\sqrt{n_j}} & \mbox{if $x_i \in \C_j$ } \\
0 & \mbox{otherwise}
\end{cases} .
\end{equation}
Throughout the paper, we use the notation $M_{i\bullet}$ to denote
the $i$th row of a matrix $M$, and $M_{\bullet j}$ denotes its $j$th column.
Our next result shows that the optimization problem \eqref{eq:minimize}
is NP-hard since
it is a quadratically constrained quadratic program (QCQP).

@@

Based on Proposition~\ref{th:qcqp2}, to group data
$\mathbb{X} = \{ x_1,\dotsc,x_n \}$
into  $k$ clusters we first compute the Gram matrix
$G$ and then 
solve the optimization problem \eqref{eq:qcqp2} for $\Zt \in
\mathbb{R}^{n\times k}$. The $i$th row
of $\Zt$ will contain a single nonzero element in some $j$th column,
indicating that $x_i \in \C_j$. 
Problem \eqref{eq:qcqp2} is NP-hard and there
are few methods
available to solve it directly,
which is computational prohibitive even for small datasets.
However, one can find approximate solutions by relaxing some 
of the constraints, or obtaining a relaxed SDP version of it.
For instance, the relaxed problem
\begin{equation}
\max_{Y} \Tr \left( Y^\top G \, Y \right) \quad \mbox{s.t. $Y^\top Y = I$}
\end{equation}
has a well-known closed form solution $Y^\star = U R$, where the
columns of $U$ contain the leading $k$ eigenvectors of $G$ corresponding
to the $k$ largest eigenvalues $\lambda_1\ge \lambda_2\ge\dotsc\ge\lambda_k$, 
and
$R \in \mathbb{R}^{k\times k}$ is an arbitrary orthogonal matrix. 
The resulting
optimal objective function is given by
$\max \Tr \left( {Y^\star}^\top G \, Y^\star \right)  = 
\sum_{i=1}^k \lambda_i$. One might then normalize and threshold the rows
of $Y^\star$, or better, following \cite{NgJordan} one can normalize the
rows of $Y^\star$ and apply standard $k$-means on this matrix where each
row is considered as a data point.
This is the procedure done in spectral clustering on a graph Laplacian
matrix obtained through a similarity matrix.
However, computing eigenvectors of very large matrices
can be prohibitively expensive 
and usually iterative methods are preferred. Moreover,
this procedure is only guaranteed to converge if the normalized $Y^\star$ is
positive semidefinite.
In Section~\ref{sec:algo} we will propose an iterative method to directly 
find approximate solutions to \eqref{eq:qcqp2} that is guaranteed to
converge and does not require that the similarities $\kk(x_i,x_j)$ be
nonnegative.

The clustering problem \eqref{eq:qcqp2} based on energy statistics
is valid for data living in an \emph{arbitrary} space of negative type where
a semimetric $\rho$, and thus the kernel \eqref{eq:kernel_semimetric}, are
assumed to be known. This clustering method is
\emph{nonparametric} since it does not make any assumptions
about  the distribution of the data, 
contrary to $k$-means and gaussian mixture models (GMM), for example.
%Moreover, this approach \emph{does not} require the concept of the 
%\emph{cluster mean},
%which can be ill-defined for some types of data, such as images for
%instance. Since no mean is involved, the method should also be 
%robust to outliers.
%This method is quite general and makes very few
%assumptions about the data. 
If one uses the standard energy distance
defined in \eqref{eq:energy} 
we have $\rho(x,y) = \| x - y\|$ and this fixes the kernel
through \eqref{eq:kernel_semimetric}. In the same way, 
for data living in a more general
metric space $(\mathcal{X}, \rho)$ the corresponding semimetric $\rho$ fixes
the kernel.
In practice, however,
the clustering quality strongly depend on the choice of a suitable
$\rho$ which measures the similarity between data points,
and is equivalent to choosing an appropriate kernel.
The energy distance \eqref{eq:energy2} fixes a family of choices
$\rho(x,y) = \| x-y\|^\alpha$ with $0 < \alpha\le 2$. 
Nevertheless, if prior information
is available one may choose a more appropriate $\rho$.

\subsection*{Relation to Kernel $\bm{k}$-Means}
One may wonder how energy statistics clustering 
relates to the well-known kernel $k$-means problem%
\footnote{When we refer to kernel $k$-means problem we mean specifically 
the optimization problem \eqref{eq:kernel_kmeans}, which should not be 
confused with kernel $k$-means algorithm that is just one possible recipe 
to solve \eqref{eq:kernel_kmeans}.
} 
which is extensively used in machine learning.
%and was proposed as a kernelized version of $k$-means.
For a positive semidefinite Gram matrix $G$, as defined in
\eqref{eq:kernel_matrix},
there exists a map
$\varphi: \mathcal{X} \to \HH_\kk$ such that
$\kk(x,y) = \varphi(x)^\top \varphi(y)$. The kernel $k$-means optimization
problem,
in feature space,
is defined by
\begin{equation}
\label{eq:kernel_kmeans}
\min_{\C_1,\dotsc,\C_k}\bigg\{ 
J(\C_1,\dots,\C_k) \equiv  \sum_{j=1}^k
\sum_{x \in \C_j} \| \varphi(x) - \varphi(\mu_j) \|^2
\bigg\}
\end{equation}
where $\mu_j = \tfrac{1}{n_j} \sum_{x \in \C_j} x$ is the  mean of cluster
$\C_j$ in the ambient space. Notice that the above objective function
is strongly tied to the idea of minimizing distances between points
and cluster centers, which arises from $k$-means objective function.
It was shown \cite{Dhillon2,Dhillon} 
that problem \eqref{eq:kernel_kmeans} 
is equivalent to a QCQP in the same form as
\eqref{eq:qcqp2}. The next result makes this explicit, showing that
\eqref{eq:minimize} and \eqref{eq:kernel_kmeans} are actually equivalent.

@@

\begin{proposition}
\label{th:qcqp3}
The weighted version of energy statistics clustering given by
problem \eqref{eq:minimize2} is equivalent to
\begin{equation}
\label{eq:qcqp3}
\max_H \Tr \left\{ H^\top (\mathcal{W}^{1/2} G \mathcal{W}^{1/2}) H  \right\}
\qquad \mbox{s.t. $H \ge 0$, $H^\top H = I$, $H H^\top \om =
\om$,}
\end{equation}
where $G$ is the Gram matrix \eqref{eq:kernel_matrix} and the other quantities
are defined in 
\eqref{eq:weighted_matrices}.
\end{proposition}
\begin{proof}
Replacing \eqref{eq:gen_kernel} and eliminating the global terms which 
do not contribute, the optimization problem \eqref{eq:minimize2}
becomes 
\begin{equation}
\max_{\C_1,\dotsc,\C_k} \sum_{j=1}^k \dfrac{1}{s_j}
\sum_{x\in\C_j}\sum_{y\in\C_j} w(x)w(y) \kk(x,y) . 
\end{equation}
This 
objective function can be written as
\begin{equation}
\begin{split}
\sum_{j=1}^k \dfrac{1}{s_j} 
\sum_{p=1}^n \sum_{q=1}^n 
w_p w_q Z_{pj} Z_{qj} G_{pq} &= 
\sum_{j=1}^k 
\sum_{p=1}^n \sum_{q=1}^n 
\dfrac{Z^\top_{jp}\sqrt{w_p}}{\sqrt{s_j}} w_p^{1/2} G_{pq} w_q^{1/2} 
\dfrac{\sqrt{w_q} Z_{qj}}{\sqrt{s_j}} \\
&= 
\sum_{j=1}^k \left(H^\top \mathcal{W}^{1/2} G \mathcal{W}^{1/2} H\right)_{jj}
\\
&= \Tr\left( H^\top \mathcal{W}^{1/2} G \mathcal{W}^{1/2} H  \right).
\end{split}
\end{equation}
To obtain the constraints, note that $H_{ij} \ge 0$ by definition, and
\begin{equation}
(H^\top H)_{ij} = \sum_{\ell=1}^n 
Y_{\ell i} \mathcal{W}_{\ell \ell} Y_{\ell j } = 
\dfrac{1}{\sqrt{s_i}\sqrt{s_j}} \sum_{\ell=1}^n w_\ell Z_{\ell i} Z_{\ell j}
= \dfrac{\delta_{ij}}{s_i} \sum_{\ell=1}^n w_\ell Z_{\ell i} = \delta_{ij},
\end{equation}
therefore $H^\top H = I$. This is a constraint on the rows of $H$.
To obtain a condition on its columns
observe that
\begin{equation}
\left(H^\top H\right)_{pq} = \sqrt{w_p w_q}\sum_{j=1}^k \dfrac{Z_{pj}
Z_{qj}}{s_j} = \begin{cases}
\dfrac{\sqrt{w_p w_q}}{s_i} & \mbox{if both $x_p,x_q \in \C_i$} \\
0 & \mbox{otherwise}.
\end{cases}
\end{equation}
Therefore, $(H^\top H \mathcal{W}^{1/2})_{pq} = \sqrt{w_p} \, w_q s_i^{-1}$
if both points $x_p$ and $x_q$ belong to the same cluster, which
we denote by $\C_i$ for some $i\in\{1,\dotsc,k\}$, and 
$(H^\top H \mathcal{W}^{1/2})_{pq} = 0 $ otherwise. Thus, the $p$th
line of this matrix is nonzero only on entries corresponding to points
that are in the same cluster as $x_p$. If we sum over the columns of this
line we obtain $\sqrt{w_p} s_i^{-1} \sum_{q=1}^n w_q Z_{qi} = \sqrt{w_p}$,
or equivalently $H H^\top \mathcal{W}^{1/2} \e = \mathcal{W}^{1/2} \e$,
which gives the constraint $H H^\top \om = \om$.
\end{proof}

@@

The relation between kernel $k$-means and graph partitioning problems
is known \cite{Dhillon2,Dhillon}. For conciseness, we repeat a similar 
analysis due to the relation of these problems to
energy statistics and RKHS, which may provide a different perspective.
